Are there unconscious visual images in aphantasia? Development of an implicit priming paradigm

### 2024-09-22 notebooks/03-manuscript-revised.qmd
Done, ready to wrap-up
- Manuscript fully transferred and highlighted, figures renamed, etc.
- Response letter and manuscript synchronized
- Bayesian integrated into main modelling notebook
- All notebooks rendered and reproducible
- All docx produced
- Useless material cleaned
- Ready to welcome the last manuscript changes and wrap-up

---
title: "Are there unconscious visual images in aphantasia? Development of an implicit priming paradigm"
abstract: For some people the experience of visual imagery is lacking, a condition recently referred to as aphantasia. So far, most of the studies on aphantasia rely on subjective reports, leaving the question of whether mental images can exist without reaching consciousness unresolved. In the present study, the formation of mental images was estimated in individuals with aphantasia without explicitly asking them to generate mental images. 151 Participants performed an implicit priming task where a probe is assumed to automatically reactivate a mental image. An explicit priming task, where participants were explicitly required to form a mental image after a probe, served as a control task. While control participants showed a priming effect in both the implicit and explicit tasks, aphantasics did not show any priming effects. These results suggest that aphantasia relies on a genuine inability to generate mental images rather than on a deficit in accessing these images consciously. Our priming paradigm might be a promising tool for characterizing mental images without relying on participant introspection.
keywords: 
  - Aphantasia 
  - visual imagery 
  - sensory priming
author: 
  - name: Rudy Purkart
    orcid: 0000-0002-5491-9958
    equal-contributor: true
    affiliations:
    - ref: emc
    - id: criugm
      name: Research Center of the Institut Universitaire de Gériatrie de Montréal (CRIUGM)
      city: Montreal
      state: Quebec
      country: Canada
      url: https://www.criugm.qc.ca
  - name: Maël Delem
    orcid: 0009-0005-8518-1991
    equal-contributor: true
    affiliations:
    - id: emc
      name: Study of Cognitive Mechanisms (EMC) Laboratory
      city: Lyon
      country: France
      url: https://emc.univ-lyon2.fr
  - name: Eddy Cavalli
    orcid: 0000-0003-3944-1973
    affiliations:
    - ref: emc
  - name: Virginie Ranson
    affiliations:
    - ref: emc
  - name: Charlotte Andrey
    affiliations:
    - ref: emc
  - name: Rémy Versace
    orcid: 0000-0002-4219-6617
    affiliations:
    - ref: emc
  - name: Gaën Plancher
    orcid: 0000-0002-0178-6207
    affiliations:
    - ref: emc
    - id: iuf
      name: Institut Universitaire de France (IUF)
      city: Paris
      country: France
      url: https://www.iufrance.fr
    email: gaen.plancher@univ-lyon2.fr
    corresponding: true
format:
  docx: default
---

@@

Close your eyes and try to mentally visualize your breakfast table as you sat down to it this morning. Most people would report having a clear image of the scene in their "*mind's eye*": this experience is referred to as visual imagery, commonly defined as the experience of visual sensory information without a direct external stimulus [@pearson_human_2019]. Nonetheless, there are great differences between individuals concerning the vividness of these images, some people even declaring that they have no visual mental images at all. This extreme phenomenon, although already observed in the $19^{th}$ century [@galton1880] and noted in various studies since then [e.g., @farahCaseStudyMental1988; @fawConflictingIntuitionsMay2009; @marksVisualImageryDifferences1973; @mckelvieVVIQPsychometricTest1995; @paivioImageryAbilityVisual1971; @sheehanUnderstandingVariabilityImagery1987; @slatterAlphaRhythmsMental1960], received renewed attention only nine years ago, when @zemanLivesImageryCongenital2015 published a study that has since been highly cited, coining the term "*aphantasia*" to refer to "reduced or absent voluntary imagery".

@@

Some researchers doubt that mental imagery could be completely absent, and have hypothesized that aphantasics who report having no conscious imagery at all might still have *unconscious* mental imagery [@nanay_unconscious_2020]. As this phenomenon is difficult to identify, little is known about unconscious mental imagery in aphantasia. A recent study by @liuProbingUnimaginableImpact2023 used a behavioural task to assess visual imagery in various domains (the French Perception-Imagination Battery) in aphantasics and controls. The task consisted of mentally comparing pairs of items based on various features (e.g. "beaver" - "fox": which is the *longest*?). They found no differences in accuracy between the groups, but slower RTs and lower confidence in the aphantasics' answers, and argued that this result was consistent with aphantasics having the visual images required to succeed at the task , albeit without the knowledge of "how they did it". Similarly, in a task that supposedly required visual imagery to verify whether a target dot was inside the boundaries of a previously presented geometric shape, @jacobsVisualWorkingMemory2018 found no difference in accuracy between an aphantasic participant and controls. As proposed by @nanay_unconscious_2020, this result could be interpreted as reflecting an unconscious comparison of a visual image with the stimuli perceived by the aphantasic participant. Although on a N = 1 case study, this result may still suggest the existence of this unconscious form of imagery in aphantasia. However, due to the potential conscious processes at play in these behavioural tasks (with or without sensory imagery), neither study could completely rule out the hypothesis of aphantasics using strategies other than visual imagery (e.g. semantic processing, spatial imagery) to solve the tasks. This possibility, often raised in discussions about aphantasia [e.g., @jacobsVisualWorkingMemory2018; @knightMemoryImageryNo2022; @liuProbingUnimaginableImpact2023; @monzelImagineYouWill2021], prevents from firmly concluding from these results regarding the existence of unconscious mental images.

On the other hand, several studies that sought to develop objective measures assessing visual imagery found consistent behavioural and physiological differences between aphantasics and controls on imagery tasks, thus challenging Nanay's (2020) hypothesis of unconscious mental imagery in aphantasia [see also @blomkvistAphantasiaSearchTheory2022]. Aphantasics have been shown to have a reduced skin conductance response to frightening scenarios [@wickenCriticalRoleMental2021], a reduced automatic pupil dilatation in reaction to imagined bright stimuli [@kayPupillaryLightResponse2022], or a reduced priming by visual imagery [@keoghBlindMindNo2018], suggesting that aphantasics are truly unable to produce mental images. The latter priming paradigm developed by [@keoghBlindMindNo2018] is particularly often cited as a promising and relatively undemanding task for objectively identifying aphantasia, and has already been used for this purpose in several subsequent studies [e.g., @keoghAttentionDrivenPhantom2020; @changImagelessImageryAphantasia2023]. In their initial study, they investigated visual imagery in aphantasics and controls using a binocular rivalry paradigm: in this task, participants were cued either with the letter “R” (for red) or the letter “G” (for green) and had to imagine one of two images, respectively a red-horizontal Gabor or a green-vertical Gabor. After rating the vividness of their mental image, they were presented with both Gabors simultaneously, one in the left eye, the other in the right, and asked to say which colour they had seen first. Their results showed that the mental visualization of the Gabors influenced the colour seen in the binocular rivalry task for control participants, but not for aphantasics. The authors interpreted this absence of priming in aphantasia as a real inability to generate mental images, and not just as a lack of metacognition skills.

While these conclusions are convincing about conscious imagery, they cannot yet rule out the hypothesis of unconscious mental imagery. A first caveat in this study is that self-diagnosed aphantasics were explicitly asked to *voluntarily* form mental images and rate their vividness during the task. This aspect of their paradigm may have skewed the results from the start, as aphantasics were asked to do something they knew (or believed) they could not do in the first place. It is possible, then, that participants did not fully engage with the task - or did not perform it correctly - because they firmly believed that they would fail to comply with the instructions, due to the awareness of their aphantasia [see @cabbaiInvestigatingRelationshipsTrait2023 for evidence on demand biases in aphantasia]. Secondly, by using explicit priming, their study could not account for potential unconscious mental imagery, which is typically investigated with implicit priming tasks. Consequently, the binocular rivalry paradigm developed by [@keoghBlindMindNo2018; like other objective measures based on explicit instructions to use mental imagery, e.g., @kayPupillaryLightResponse2022; @miltonBehavioralNeuralSignatures2021] cannot exclude the possible existence of unconscious mental images in aphantasia. We aimed to fill this gap by designing an implicit priming task that would allow us to study unconscious mental images in aphantasia.

@@

To create an implicit task, the paradigm was divided into two parts: a prior association phase, and the implicit task itself. The association phase asked the participants to indicate the colour of a red horizontal Gabor or a blue vertical Gabor to implicitly memorize the colour-orientation association. In the subsequent implicit task, a red or blue coloured circle was presented as a prime before a target, which was a Gabor either congruent or incongruent with the colour seen (see Figures \ref{fig:protocol_asso} and \ref{fig:protocol_implicit} in the Procedure section). As previous works of our team have shown [e.g., @brunelWhenSeeingDog2013; @reyMaskWhoWasn2015; @reyWhenReactivatedVisual2018], the prime is assumed to reactivate an unconscious mental image of the associated Gabor automatically. Participants were then asked to indicate the orientation of the target. An explicit task was added for control and comparison where participants were overtly asked to produce a mental image of the Gabors before the targets. In the explicit task, like in Keogh and Pearson’s study, participants were asked to imagine one of the previous Gabors by presenting a letter (R or B) as a cue. A Gabor congruent or incongruent with the colour was then presented as a target, and participants were asked to indicate the orientation of the lines of the target.

Shorter RTs in the congruent trials compared to the incongruent ones would therefore reflect an influence of unconscious mental images, helping participants to respond faster. Moreover, to ensure that any priming observed would be the consequence of mental imagery (as opposed to semantic priming, for example), in half of the trials the targets were presented in black and white rather in than colour. If aphantasics have only difficulties with conscious mental imagery, a priming effect should be observed for both groups in the implicit task, but not in the explicit one. If aphantasics have difficulties with both conscious and unconscious mental imagery, no priming effect should be observed for the aphantasia group neither in the implicit nor in the explicit task, as opposed to the control group.

In addition, we carried out a second analysis by defining "finer-grained" VVIQ sub-groups, with a subset of our sample composed solely of aphantasics with a minimal VVIQ score of 16, people with low imagery (VVIQ between 17 and 32), hereinafter called "hypophantasics" [using the terminology from @reederNonvisualSpatialStrategies2024], and controls (our sample included only two participants with VVIQ $>$ 75, so group analyses with hyperphantasics were not possible). This analysis aimed at answering a frequent interrogation in aphantasia literature about subgroups in aphantasia with differing characteristics: several authors pointed out that the fact of reporting *completely absent* imagery could be qualitatively very distinct from having only *vague* images [@blomkvistDefiningDiagnosingAphantasia2023; @dancePrevalenceAphantasiaImagery2022; @murakiInsightsEmbodiedCognition2023; @liuProbingUnimaginableImpact2023]. The object of our study, unconscious mental images, is a difficult phenomenon to reach, so this additional analysis could be crucial to judge the effects of our paradigm.

@@

The French version of the VVIQ [VVIQ-F, @santarpiaEvaluerVivaciteImages2008; adapted from @marksVividnessVisualImagery1973] used in this study was adapted in a Google Form version. It consists of sixteen items, each asking participants to imagine a particular scene and rate the vividness of their mental imagery using a Likert scale ranging from 1 ("No image at all, you only know that you are thinking about the object”) to 5 ("Perfectly clear and vivid as if it [were normal]{bg-colour="#FFFF00"} vision").

@@

![***Associative task.*** A fixation cross (500 ms) is followed by one of the two coloured Gabor. Participants had to indicate the colour of the Gabor by pressing the corresponding key (either R for Red or B for Blue), without time constraint. Each Gabor was presented 50 times.](../figures/protocol-implicit.png){#fig-protocol_asso}

Participants began with the implicit priming task. This task began with an associative phase (@fig-protocol_asso) in which a fixation cross (500 ms) was followed by one of the two coloured Gabor, and participants had to indicate the colour of the Gabor by pressing the corresponding key (either R for Red or B for Blue), without time constraint. Each Gabor was presented 50 times. After the association phase, participants started the implicit priming test phase (@fig-protocol_implicit) in which a fixation cross (500 ms) was followed by a cue (150 ms) that was either a red or a blue circle, and then depending on the condition by either one of the two coloured Gabor or by one of the uncoloured Gabor, as a target. Participants had to indicate the orientation of the Gabor’s lines by pressing the corresponding key (either H for horizontal, or V for vertical), using only their dominant hand. In the congruent condition, the cue has the same colour as the target Gabor (coloured condition) or primed the Gabor with the same line orientation as the black target Gabor (uncoloured condition), contrary to the non-congruent condition. There were 16 trials per congruence and colour pairs, amounting to 64 trials total for the task.

![***Implicit priming task.*** A fixation cross (500 ms) is followed by a cue (150 ms) that was either a red or a blue circle, and then by one of the two coloured Gabor, or by one of the uncoloured Gabor, as a target. Participants had to indicate the orientation of the Gabor’s lines by pressing the corresponding key (either H for horizontal, or V for vertical), using only their dominant hand.](../figures/protocol-implicit.png){#fig-protocol_implicit}

After the implicit priming task and a short break, participants transitioned to the explicit task (@fig-protocol_explicit) in which a fixation cross (500 ms) was followed by a letter as a cue (1500 ms) that was either the letter R (for red) or B (for blue). In response to the cue, participants were asked to form a mental imagery of the corresponding Gabor (a red with horizontal lines or a blue with vertical lines) and to keep that image in mind during 3000 ms while fixing the centre of a blank screen. Then, a fixation cross (500 ms) was presented followed by one of the two coloured Gabor, or by one of the uncoloured Gabor, and participants had to indicate the orientation of the Gabor’s lines by pressing the corresponding key (either H for horizontal, or V for vertical), using only their dominant hand. In the congruent condition, the cue designated the same colour as the target Gabor (coloured condition) or primed the Gabor with the same line orientation as the black target Gabor (uncoloured condition), contrary to the non-congruent condition. Likewise, there were 16 trials per congruence and colour pairs, amounting to 64 trials total for the task. It is worth mentioning that participants had practice trials at the beginning of each task and phase. Moreover, the implicit priming task was divided into two blocks, each beginning with the association phase, and ending with the implicit priming phase, to preserve the association between the colour of the Gabor and the orientation of its lines.

![***Explicit priming task.*** A fixation cross (500 ms) was followed by a letter as a cue (1500 ms) that was either the letter R (for red) or B (for blue). In response to the cue, participants were asked to form a mental imagery of the corresponding Gabor (a red with horizontal lines or a blue with vertical lines) and to keep that imagery in mind during 3000 ms while fixing the centre of a blank screen. Then, a fixation cross (500 ms) was presented followed by one of the two coloured Gabor, or by one of the uncoloured Gabor, and participants had to indicate the orientation of the Gabor’s lines by pressing the corresponding key (either H for horizontal, or V for vertical), using only their dominant hand.](../figures/protocol-explicit.png){#fig-protocol_explicit}

@@

The scores of the four questionnaires were modelled with Generalized Linear Models (GLMs) to accommodate the non-normal distributions of several questionnaires (mostly bi-modal distributions due to our two groups) and to control for the influence of age on the group differences. The GLMs were fitted with Gamma distributions, chosen as the best one against Gaussian and inverse Gaussian distributions fitted and compared using the Akaike Information Criterion (AIC) and Bayesian Information Criterion (BIC). The models included the *Group* as a categorical predictor and the *Age* as a continuous covariate. For group differences, we report estimated means, their standard errors, and two-tailed *p*-values of marginal contrasts between means, computed using a Wald *t*-distribution approximation.

>>>>

---
title: "Are there unconscious visual images in aphantasia? Development of an implicit priming paradigm"
abstract: For some people the experience of visual imagery is lacking, a condition recently referred to as aphantasia. So far, most of the studies on aphantasia rely on subjective reports, leaving the question of whether mental images can exist without reaching consciousness unresolved. In the present study, the formation of mental images was estimated in individuals with aphantasia without explicitly asking them to generate mental images. 151 Participants performed an implicit priming task where a probe is assumed to automatically reactivate a mental image. An explicit priming task, where participants were explicitly required to form a mental image after a probe, served as a control task. While control participants showed a priming effect in both the implicit and explicit tasks, aphantasics did not show any priming effects. These results suggest that aphantasia relies on a genuine inability to generate mental images rather than on a deficit in accessing these images consciously. Our priming paradigm might be a promising tool for characterizing mental images without relying on participant introspection.
keywords: 
  - Aphantasia 
  - visual imagery 
  - sensory priming
author: 
  - name: Rudy Purkart
    orcid: 0000-0002-5491-9958
    equal-contributor: true
    affiliations:
    - ref: emc
    - id: criugm
      name: Research Center of the Institut Universitaire de Gériatrie de Montréal (CRIUGM)
      city: Montreal
      state: Quebec
      country: Canada
      url: https://www.criugm.qc.ca
  - name: Maël Delem
    orcid: 0009-0005-8518-1991
    equal-contributor: true
    affiliations:
    - id: emc
      name: Study of Cognitive Mechanisms (EMC) Laboratory
      city: Lyon
      country: France
      url: https://emc.univ-lyon2.fr
  - name: Eddy Cavalli
    orcid: 0000-0003-3944-1973
    affiliations:
    - ref: emc
  - name: Virginie Ranson
    affiliations:
    - ref: emc
  - name: Charlotte Andrey
    affiliations:
    - ref: emc
  - name: Rémy Versace
    orcid: 0000-0002-4219-6617
    affiliations:
    - ref: emc
  - name: Gaën Plancher
    orcid: 0000-0002-0178-6207
    affiliations:
    - ref: emc
    - id: iuf
      name: Institut Universitaire de France (IUF)
      city: Paris
      country: France
      url: https://www.iufrance.fr
    email: gaen.plancher@univ-lyon2.fr
    corresponding: true
format:
  docx: 
    reference-doc: "../utils/docx/custom-reference-doc-numbered.docx"
---

::: {.content-hidden when-format="pdf"}
**Keywords**: Aphantasia, visual imagery, sensory priming
:::

<!-- Different versions of the manuscript to build: -->

<!--  1. Full manuscript + title + abstract (automatically generated) -->

<!--  2. Manuscript anonymised (no title page, remove author contributions) -->

<!--  3. Title page -->

<!--  4. Title page with authors only -->

<!--  5. References only -->

@@

Close your eyes and try to mentally visualize your breakfast table as you sat down to it this morning. Most people would report having a clear image of the scene in their "*mind's eye*": this experience is referred to as visual imagery, commonly defined as the experience of visual sensory information without a direct external stimulus [@pearson_human_2019]. Nonetheless, there are great differences between individuals concerning the vividness of these images, some people even declaring that they have no visual mental images at all. This extreme phenomenon, although already observed in the $19^{th}$ century [@galton1880] and noted in various studies since then [e.g., @farahCaseStudyMental1988; @fawConflictingIntuitionsMay2009; @marksVisualImageryDifferences1973; @mckelvieVVIQPsychometricTest1995; @paivioImageryAbilityVisual1971; @sheehanUnderstandingVariabilityImagery1987; @slatterAlphaRhythmsMental1960], received renewed attention only nine years ago, when @zemanLivesImageryCongenital2015 published [a study that has since been highly cited]{bg-colour="#FFFF00"}, coining the term "*aphantasia*" to refer to "[reduced or absent voluntary imagery]{bg-colour="#FFFF00"}".

@@

Some researchers doubt that mental imagery could be completely absent, and have hypothesized that aphantasics who report having no conscious imagery at all might still have *unconscious* mental imagery [@nanay_unconscious_2020]. As this phenomenon is difficult to identify, little is known about unconscious mental imagery in aphantasia. A recent study by @liuProbingUnimaginableImpact2023 used a behavioural task to assess visual imagery in various domains (the French Perception-Imagination Battery) in aphantasics and controls. The task consisted of mentally comparing pairs of items based on various features (e.g. "beaver" - "fox": which is the *longest*?). They found no differences in accuracy between the groups, but slower RTs and lower confidence in the aphantasics' answers, and argued that this result was consistent with aphantasics having the visual images required [to succeed at the task, albeit without the knowledge of "how they did it"]{bg-colour="#FFFF00"}. Similarly, in a task that supposedly required visual imagery to verify whether a target dot was inside the boundaries of a previously presented geometric shape, @jacobsVisualWorkingMemory2018 found no difference in accuracy between an aphantasic participant and controls. As proposed by @nanay_unconscious_2020, this result could be interpreted as reflecting an unconscious comparison of a visual image with the stimuli perceived by the aphantasic participant. [Although on a N = 1 case study, this result may still suggest]{bg-colour="#FFFF00"} the existence of this unconscious form of imagery in aphantasia. However, due to the potential conscious processes at play in these behavioural tasks (with or without sensory imagery), neither study could completely rule out the hypothesis of aphantasics using strategies other than visual imagery (e.g. semantic processing, spatial imagery) to solve the tasks. This possibility, often raised in discussions about aphantasia [[e.g., @jacobsVisualWorkingMemory2018; @knightMemoryImageryNo2022; @liuProbingUnimaginableImpact2023; @monzelImagineYouWill2021]{bg-colour="#FFFF00"}], prevents from firmly concluding from these results regarding the existence of unconscious mental images.

On the other hand, several studies that sought to develop objective measures assessing visual imagery found [consistent behavioural and physiological differences between aphantasics and controls on imagery tasks, thus challenging Nanay's (2020) hypothesis of unconscious mental imagery in aphantasia [see also @blomkvistAphantasiaSearchTheory2022].]{bg-colour="#FFFF00"} Aphantasics have been shown to have a [reduced]{bg-colour="#FFFF00"} skin conductance response to frightening scenarios [@wickenCriticalRoleMental2021], a reduced automatic pupil dilatation in reaction to imagined bright stimuli [@kayPupillaryLightResponse2022], or a reduced priming by visual imagery [@keoghBlindMindNo2018], suggesting that aphantasics are truly unable to produce mental images. The latter priming paradigm developed by [@keoghBlindMindNo2018] is particularly often cited as a promising and relatively undemanding task for objectively identifying aphantasia, and has already been used for this purpose in several subsequent studies [e.g., @keoghAttentionDrivenPhantom2020; @changImagelessImageryAphantasia2023]. In their initial study, they investigated visual imagery in aphantasics and controls using a binocular rivalry paradigm: in this task, participants were cued either with the letter “R” (for red) or the letter “G” (for green) and had to imagine one of two images, respectively a red-horizontal Gabor or a green-vertical Gabor. After rating the vividness of their mental image, they were presented with both Gabors simultaneously, one in the left eye, the other in the right, and asked to say which colour they had seen first. Their results showed that the mental visualization of the Gabors influenced the colour seen in the binocular rivalry task for control participants, but not for aphantasics. The authors interpreted this absence of priming in aphantasia as a real inability to generate mental images, and not just as a lack of metacognition skills.

While these conclusions are convincing about [conscious]{bg-colour="#FFFF00"} imagery, they cannot yet rule out the hypothesis of unconscious mental imagery. A first caveat in this study is that self-diagnosed aphantasics were explicitly asked to *voluntarily* form mental images and rate their vividness during the task. This aspect of their paradigm may have skewed the results from the start, as aphantasics were asked to do something they knew (or believed) they could not do in the first place. It is possible, then, that participants did not fully engage with the task - or did not perform it correctly - because they firmly believed that they would fail to comply with the instructions, due to the awareness of their aphantasia [see @cabbaiInvestigatingRelationshipsTrait2023 for evidence on demand biases in aphantasia]. Secondly, by using explicit priming, their study could not account for potential unconscious mental imagery, which is typically investigated with implicit priming tasks. Consequently, the binocular rivalry paradigm developed by [@keoghBlindMindNo2018; like other objective measures based on explicit instructions to use mental imagery, e.g., @kayPupillaryLightResponse2022; @miltonBehavioralNeuralSignatures2021] cannot exclude the possible existence of [unconscious]{bg-colour="#FFFF00"} mental images in aphantasia. We aimed to fill this gap by designing an implicit priming task that would allow us to study unconscious mental images in aphantasia.

@@

To create an implicit task, the paradigm was divided into two parts: a prior association phase, and the implicit task itself. The association phase asked the participants to indicate the colour of a red horizontal Gabor or a blue vertical Gabor to implicitly memorize the colour-orientation association. In the subsequent implicit task, a red or blue coloured circle was presented as a prime before a target, which was a Gabor either congruent or incongruent with the colour seen (see @fig-protocol-asso and @fig-protocol-implicit in the Procedure section). As previous works of our team have shown [e.g., @brunelWhenSeeingDog2013; @reyMaskWhoWasn2015; @reyWhenReactivatedVisual2018], the prime is assumed to reactivate an unconscious mental image of the associated Gabor automatically. Participants were then asked to indicate the orientation of the target. An explicit task was added for control and comparison where participants were overtly asked to produce a mental image of the Gabors before the targets. In the explicit task, like in Keogh and Pearson’s study, participants were asked to imagine one of the previous Gabors by presenting a letter (R or B) as a cue. A Gabor congruent or incongruent with the colour was then presented as a target, and participants were asked to indicate the orientation of the lines of the target.

Shorter RTs in the congruent trials compared to the incongruent ones would therefore reflect an influence of unconscious mental images, helping participants to respond faster. Moreover, to ensure that any priming observed would be the consequence of mental imagery (as opposed to semantic priming, for example), in half of the trials the targets were presented in black and white rather in than colour. [If aphantasics have only difficulties with conscious mental imagery, a priming effect should be observed for both groups in the implicit task, but not in the explicit one. If aphantasics have difficulties with both conscious and unconscious mental imagery, no priming effect should be observed for the aphantasia group neither in the implicit nor in the explicit task, as opposed to the control group.]{bg-colour="#FFFF00"}

[In addition, we carried out a second analysis by defining "finer-grained" VVIQ sub-groups]{bg-colour="#FFFF00"}, with a subset of our sample composed solely of aphantasics with a minimal VVIQ score of 16, people with low imagery (VVIQ between 17 and 32), hereinafter called "hypophantasics" [using the terminology from @reederNonvisualSpatialStrategies2024], and controls (our sample included only two participants with VVIQ $>$ 75, so group analyses with hyperphantasics were not possible). This analysis aimed at answering a frequent interrogation in aphantasia literature about subgroups in aphantasia with differing characteristics: several authors pointed out that the fact of reporting *completely absent* imagery could be qualitatively very distinct from having only *vague* images [@blomkvistDefiningDiagnosingAphantasia2023; @dancePrevalenceAphantasiaImagery2022; @murakiInsightsEmbodiedCognition2023; @liuProbingUnimaginableImpact2023]. The object of our study, unconscious mental images, is a difficult phenomenon to reach, so this additional analysis could be crucial to judge the effects of our paradigm.

@@

The French version of the VVIQ [VVIQ-F, @santarpiaEvaluerVivaciteImages2008; adapted from @marksVividnessVisualImagery1973] used in this study was adapted in a Google Form version. It consists of sixteen items, each asking participants to imagine a particular scene and rate the vividness of their mental imagery using a Likert scale ranging from 1 ("No image at all, you only know that you are thinking about the object”) to 5 ("Perfectly clear and vivid as if it [were normal]{bg-colour="#FFFF00"} vision"). [The final score is ranging from 16 to 80.]{bg-colour="#FFFF00"}

@@

![***Associative task.*** A fixation cross (500 ms) is followed by one of the two coloured Gabor. Participants had to indicate the colour of the Gabor by pressing the corresponding key (either R for Red or B for Blue), without time constraint. Each Gabor was presented 50 times.](../figures/protocol-asso.png){#fig-protocol-asso}

Participants began with the implicit priming task. This task began with an associative phase (@fig-protocol-asso) in which a fixation cross (500 ms) was followed by one of the two coloured Gabor, and participants had to indicate the colour of the Gabor by pressing the corresponding key (either R for Red or B for Blue), without time constraint. Each Gabor was presented 50 times. After the association phase, participants started the implicit priming test phase (@fig-protocol-implicit) in which a fixation cross (500 ms) was followed by a cue (150 ms) that was either a red or a blue circle, and then depending on the condition by either one of the two coloured Gabor or by one of the uncoloured Gabor, as a target. Participants had to indicate the orientation of the Gabor’s lines by pressing the corresponding key (either H for horizontal, or V for vertical), using only their dominant hand. In the congruent condition, the cue has the same colour as the target Gabor (coloured condition) or primed the Gabor with the same line orientation as the black target Gabor (uncoloured condition), contrary to the non-congruent condition. There were 16 trials per congruence and colour pairs, amounting to 64 trials total for the task.

![***Implicit priming task.*** A fixation cross (500 ms) is followed by a cue (150 ms) that was either a red or a blue circle, and then by one of the two coloured Gabor, or by one of the uncoloured Gabor, as a target. Participants had to indicate the orientation of the Gabor’s lines by pressing the corresponding key (either H for horizontal, or V for vertical), using only their dominant hand.](../figures/protocol-implicit.png){#fig-protocol-implicit}

After the implicit priming task and a short break, participants transitioned to the explicit task (@fig-protocol-explicit) in which a fixation cross (500 ms) was followed by a letter as a cue (1500 ms) that was either the letter R (for red) or B (for blue). In response to the cue, participants were asked to form a mental imagery of the corresponding Gabor (a red with horizontal lines or a blue with vertical lines) and to keep that image in mind during 3000 ms while fixing the centre of a blank screen. Then, a fixation cross (500 ms) was presented followed by one of the two coloured Gabor, or by one of the uncoloured Gabor, and participants had to indicate the orientation of the Gabor’s lines by pressing the corresponding key (either H for horizontal, or V for vertical), using only their dominant hand. In the congruent condition, the cue designated the same colour as the target Gabor (coloured condition) or primed the Gabor with the same line orientation as the black target Gabor (uncoloured condition), contrary to the non-congruent condition. Likewise, there were 16 trials per congruence and colour pairs, amounting to 64 trials total for the task. It is worth mentioning that participants had practice trials at the beginning of each task and phase. Moreover, the implicit priming task was divided into two blocks, each beginning with the association phase, and ending with the implicit priming phase, to preserve the association between the colour of the Gabor and the orientation of its lines.

![***Explicit priming task.*** A fixation cross (500 ms) was followed by a letter as a cue (1500 ms) that was either the letter R (for red) or B (for blue). In response to the cue, participants were asked to form a mental imagery of the corresponding Gabor (a red with horizontal lines or a blue with vertical lines) and to keep that imagery in mind during 3000 ms while fixing the centre of a blank screen. Then, a fixation cross (500 ms) was presented followed by one of the two coloured Gabor, or by one of the uncoloured Gabor, and participants had to indicate the orientation of the Gabor’s lines by pressing the corresponding key (either H for horizontal, or V for vertical), using only their dominant hand.](../figures/protocol-explicit.png){#fig-protocol-explicit}

@@

The scores of the four questionnaires were modelled with linear models on ranked scores to accommodate the ordinal nature of the questionnaire data and to control for the influence of age on the group differences. The models included the *Group* as a categorical predictor and the *Age* as a continuous co-variate. For group differences, we report estimated means, their standard errors, and two-tailed *p*-values of marginal contrasts between means, computed using a Wald *t*-distribution approximation.

### Outlier detection procedure

We excluded data from five participants that exceeded 43% of errors (therefore with too few trials to analyze). Trials with RTs abnormally fast or slow ($<$ 250ms or $>$ 3s, 1% and .7% respectively) that could represent false alarms (especially given the internet-based nature of the task) were removed. Participants with an aberrant RT average exceeding the median $\pm 3 \times$ MAD were excluded for each task [MAD proved to be more robust than standard deviations for outlier detection, see @leysDetectingOutliersNot2013]. The remaining data comprised N = 9082 trials for the implicit task and N = 8824 trials for the explicit task. For the analysis of reaction times (RTs), incorrect responses were also removed (2.7% and 3.6% of trials in the implicit/explicit task respectively). The remaining RT data comprised N = 8777 trials for the implicit and N = 8603 trials for the explicit task.

### [Accuracy]{bg-colour="#FFFF00"}

[Before removing incorrect responses for the analysis of response times, we first checked for any differences in accuracy between the groups.]{bg-colour="#FFFF00"} To this end, we fitted Generalized Logistic Mixed Models to predict accuracy with the *Group* (aphantasic, control), *Congruence* condition (congruent or incongruent), and *Colour* condition (colour or uncoloured) along with all their two and three-way interactions as fixed categorical predictors, while *participants* have been included as grouping factors (i.e. "random effects"). The models were implemented in the *lme4* R package [@bates2014fitting]. Overall fixed effects were assessed with Type II Wald $\chi^{2}$ tests using the *car* R package [@R-car].

### Response times

#### Power analysis

As the experiment was conducted online, the sample size was mostly limited by the time resources of the project. Likewise, the number of trials per condition and the total amount of trials were balanced in order not to overload the Internet experiment. From these two constraints, we estimated the statistical power conferred by the sample size eventually reached and the experimental design *a priori*, i.e. based on effect sizes reported in the literature and our hypotheses. We used the *simr* package [@greenSIMRPackagePower2016] to simulate datasets reflecting the experimental design, featuring expected patterns of means and variance, and fitting generalized linear mixed models on them.

Common effect sizes reported in the literature and intra-individual trial-to-trial variability in RTs in perceptual discrimination tasks have been used to simulate data and estimate the statistical power to detect effects tied to experimental conditions [for a similar procedure, see @fucciReadyHelpNo2023]. The main effect of interest chosen was the interaction between Group and Congruence, where we hypothesized a reduction in RTs for the control group in the congruent condition that would not be present in aphantasics. Simulations have shown that the statistical power reached 80% even with little effect sizes nearing 24ms, and exceeded 90% when the effect size went up and above 30ms. Details of the procedure can be found in an extended analysis report on OSF (<https://osf.io/635dv/?view_only=72898c1e036c456b97e688629563a47f>).

#### Generalized Linear Mixed Models

To account for the non-normal, positively skewed distributions of the RTs, we fitted Generalized Linear Mixed Models (GLMMs) with inverse Gaussian distributions, as recommended by @loTransformNotTransform2015. The models included the *Group* (aphantasic, control), *Congruence* condition (congruent or incongruent), and *Colour* condition (colour or uncoloured) along with all their two and three-way interactions as fixed categorical predictors, while *participants* have been included as grouping factors (i.e. "random effects"). Overall fixed effects were assessed with Type II Wald $\chi^{2}$ tests.

Due to the way that variance is partitioned in linear mixed models [@rightsQuantifyingExplainedVariance2019], there does not exist an agreed-upon way to calculate standard effect sizes for individual terms such as main effects or interactions in these models. Thus, in line with general recommendations on how to report effect sizes [e.g., @pekReportingEffectSizes2018], we report and analyse unstandardised effect sizes for post-hoc tests in the form of estimated marginal contrasts in milliseconds (i.e. differences in model-estimated marginal means, hereinafter denoted $\Delta$) where appropriate.

#### [Bayesian Modelling]{bg-colour="#FFFF00"}

To complement the frequentist approach, Bayesian hypothesis testing was also used to quantify the evidence in favour or against a congruence effect for each group. To this end, we fitted Bayesian multilevel models (analogous to the frequentist GLMMs presented above) separately for each group to predict RTs with the *Congruence* condition (congruent or incongruent), *Colour* condition (colour or uncoloured) and their two-way interaction as fixed categorical predictors, along with *participants* as grouping factors. Models were implemented in the *brms* R package [@burknerBrmsPackageBayesian2017]. The evidence ratio of interest was then computed as the Bayes Factor between the hypothesis of a congruence effect against the null hypothesis. We report the $BF_{01}$ in favour of the null hypothesis of an absence of congruence effect for a given group alongside the contrasts in the *Group* $\times$ *Congruence* interaction. The proposed interpretations of the strength of the evidence from Bayes Factor values were based on Jeffrey's scale thresholds [see e.g., @kassBayesFactors1995].

### [Finer-grained VVIQ groups]{bg-colour="#FFFF00"}

Additional analyses were conducted with more refined groups: the aphantasic group was restricted to participants scoring at the lowest point on the VVIQ (VVIQ = 16, N = 46), an "hypophantasic" group was created with participants scoring between 17 and 32 (N = 37), and the control group was restricted to participants scoring between 33 and 74 (N = 54). Our sample included only two "hyperphantasics" (VVIQ $\geq$ 75). These two participants were therefore removed from the analyses, as it was shown that hyperphantasics constitute a group with significant differences from typical imagers [see e.g., @miltonBehavioralNeuralSignatures2021; @zemanPhantasiaPsychologicalSignificance2020], but there were not enough of them in the present study to create such a group. Similar analyses to those carried out for the two initial groups were performed.

### Correlation analyses
